---
title: "Half Pounder Genotyping Notebook"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: false
---

```{r, message=FALSE, warning=FALSE}
require(tidyverse)
require(DiagrammeR)
require(poppr)
require(genepop)
require(graph4lg)
require(related)
require(adegenet)
```


# Readme

This is an rstudio project. If you'd like to pre-rendered figures, read a summary of analysis and view code, please open the relevant html file in a browser. 


To conduct the analyses on your computer, edit or run code: clone this repository into a directory on you r local machine and open the .Rproj file in Rstudio. All data (except raw sequencing data) and analyses are available in the github repository at https://github.com/david-dayan/rogue_half_pounder.git 

__Note the section headings labeled (bad scripts) used a version of the gtseq pipeline with a bug. The steps run with the bad script were re-run and relevant sections of the notebook are labeled (corrected scripts)__

# Rationale 

ODFW has proposed using half pounder abundance as the sole index for determining sliding scale fishing regulations for winter steelhead on the Rogue as (i) half pounder abundance should integrate juvenile freshwater and early ocean conditions for steelhead regardless of whether they express the half pounder phenotype and (ii) half pounder abundance is predictive of steelhead abundance in historical dam passage counts. However, the relative proportion of winter vs summer run life histories expressed by half pounders is unknown.

This study attempts to use neutral and adaptive GTseq markers used for population and life history assignment to classify half pounders into winter or summer run steelhead life histories.

This notebook covers genotyping (bioinformatic from raw sequencing reads to called genotypes) Rogue River summer fall winter and half pounder steelhead collected in 2018 and 2019. 

# GTseq outline

```{r}
grViz("digraph flowchart {
      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle]        
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab7 [label = '@@7']
      tab8 [label = '@@8']
      # edge definitions with the node IDs
      tab1 -> tab2 [label = 'GTseq_BarcodeSplit']
      tab2 -> tab3 [label = 'GTseq_Genotyper']
      tab3 -> tab4 [label = 'GTseq_genocompile']
      tab4 -> tab5 [label = 'filtering']
      tab4 -> tab6 [label = 'GTseq_summaryfigs']
      tab3 -> tab7 [label = 'GTseq_genocompilecounts']
      tab7 -> tab5 
      tab1 -> tab8 [label = 'GTseq_seqtest']
      tab7 -> tab4 [label = 'GTseq_Omysex']
      tab6 -> tab5
      }
      [1]: 'Raw Reads'
      [2]: 'Demultiplezed fastqs'
      [3]: 'Individual genotypes'
      [4]: 'raw GTseq dataset'
      [5]: 'final filtered dataset'
      [6]: 'summary figures'
      [7]: 'read counts'
      [8]: 'marker info'
      
      ")

```

# Data Summary

## Samples

__Half Pounders__  
Samples were collected during ODFW’s Lower Rogue Seining Project in 2018 and 2019. The Lower Rogue Seining Project estimates escapement for Coho, late-run summer steelhead, half-pounder steelhead and fall chinook by beach seining near Huntley Park at approximate river mile 8, three times weekly from July through October. Half-pounder steelhead were identified as individuals with fork length 250 - 410mm and sampled in batches of up to 50 fish each day for 11 days from September 7th to October 1st 2018 totaling 384 individuals and 18 days from August 14th to September 25th 2019 totaling 331 individuals. Caudal fin clips were taken for DNA extraction, and placed in daily batch vials containing 95% ethanol. Note that due to batch collection of fin clips, that these numbers are inflated (some individuals have multiple tissue samples - identified later and filtered) 

All half-pounders are non-marked and assumed to be natural origin fish.

Also ~5% of samples represented twice in GTseq library as QAQC samples

__Adults__  
45 winter and 45 early run summer run fish.. Adult summer steelhead were sampled at the Cole River Hatchery sorting pond on 6/26/2019 and (b) adult winter steelhead were sampled at Cole Rivers Hatchery (Rogue River) and the Applegate River from adult brood stock for 2019. Finally 166 additional late-run summer fish (fall run) were sampled at the Huntley Park seine on the lower Rogue.

All late-summer run adults are unmarked and assumed to natural origin fish. Winter run fish include NOR and HOR fish. After filtering the final dataset contains 18 HOR and 22 NOR fish. Early summer run fish are all unmarked and assumed NOR.

__Sample metadata__

let's gather all the metadata into a single file:
note that intake files disagrees with other sample sizes - likely because some samples came as batch jars of fin clips, and some fin clips were from the same individual and later filtered out
```{r, message=FALSE, warning=FALSE}
#intake files
half_2018_intake <- readxl::read_xlsx("../meta_data/2018_halfpounder/OmyJC18ROGR_STHP Intake form Spread sheet.xlsx", sheet = 3)
half_2019_intake <- readxl::read_xlsx("../meta_data/2019_halfpounder/STHP Intake form Spread sheet 2019.xlsx", sheet = 1)
fall_intake <- readxl::read_xlsx("../meta_data/2019_fall/Rogue Adult Summer and Winter (06 11 20).xlsx", sheet = 1)
summer_intake <- readxl::read_xlsx("../meta_data/2019_summer/Omy Rogue2019 steelhead datasheets.xlsx", sheet = 2)
winter_intake <- readxl::read_xlsx("../meta_data/2019_winter/StW Scales for DNA (06 17 19).xlsx", sheet = 3)

#merge intakes
# first clean them up a bit to make merging easier
half_2018_intake <- half_2018_intake[,c(1,6)]
colnames(half_2018_intake)[1] <- "ID"
half_2018_intake$run <- "halfpounder"
half_2018_intake$year <- "2018"
colnames(half_2018_intake) <- c("ID", "Date", "run", "year")

half_2019_intake <- half_2019_intake[,c(1,2)]
half_2019_intake$run <- "halfpounder"
half_2019_intake$year <- "2019"
colnames(half_2019_intake) <- c("ID", "Date", "run", "year")

summer_intake <- summer_intake[,c(2,3,6)]
colnames(summer_intake) <- c("ID", "Date", "run")
summer_intake$year <- "2019"

winter_intake <- winter_intake[,c(2,3,7)]
colnames(winter_intake) <- c("ID", "Date", "run")
winter_intake$year <- "2019"

fall_intake <- subset(fall_intake, Run == "Summer")
fall_intake <- fall_intake[,c(1,3,6)]
colnames(fall_intake) <- c("ID", "Date", "run")
fall_intake$run <- "fall"
fall_intake$year <- "2019"

meta_data <- bind_rows(half_2018_intake, half_2019_intake, fall_intake, winter_intake, summer_intake)

meta_data %>%
  group_by(run, year) %>%
  tally()

```

## Sequencing QC Reports

__second lane__  
Clusters: 409,149,535  
Yield (mbase): 61,782  
% >Q30 bases: 87.13  
Average Qual: 37.18  

fastqc report available in working directory

__first lane__  
don't have this info

## data locations

Sequencing data is from multiple sequencing runs and multiple technicians/bioinformaticians. Summer Winter and 2018 half-pounders are already 

__Demultiplexed fastqs__  
/nfs1/FW_HMSC/OMalley_Lab/bohns/GTseq/OmyRogue/sample_fastqs - 2018 halfpounders
/nfs1/FW_HMSC/OMalley_Lab/bohns/GTseq/OmyRogue/baseline/sample_fastqs - 2019 winter and 2019 summer

/dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux - 2019 fall and 2019 halfpounder


__Lane 1__  
This project uses data from two GTseq libraries. One sequenced in 2019 and a second in July 2020. Need to find this info for the first lane

__Lane 2__  
/dfs/FW_HMSC/Omalley_Lab/dayan/seqdata/OmyRogue2020/Undetermined_S0_L002_R1_001.fastq.gz
GTseq runs on uncompressed files, uncompressed copy in same directory "lane2.fastq"


# Demultiplex

The first step is to demultiplex the raw sequencing file using the i5 and i7 indexes. We can actually skip this because the sequencing center already performed a demux (exact matching) for the second lane, and files are already demuxed for the first lane. Instead, we just copy the demuxed files and give them more reasonable name.  


```{r, eval=FALSE}
#generate barcode file for second lane samples
# example: Sample,PlateID,i7_name,i7_sequence,i5_name,i5_sequence
#          Sample123,P1234,i001,ACCGTA,25,CCCTAA
#          Sample321,P1234,i001,ACCGTA,26,GGCACA

#first lets get the index sequences 
index2020 <- read_tsv("metadata/index_2020.txt")
colnames(index2020) <- c("Sample","PlateID","i7_name","i7_sequence","i5_name","i5_sequence")
write_csv(index2020, "./metadata/index_2020_lane.csv")
```

```{bash, eval=FALSE}
#first move the demuxed files over
cp /nfs2/hts/illumina/200723_J00107_0245_AHHLG2BBXY_1504/L23/*Omy* /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux

cp /nfs2/hts/illumina/200723_J00107_0245_AHHLG2BBXY_1504/L23/*positive* /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux

cp /nfs2/hts/illumina/200723_J00107_0245_AHHLG2BBXY_1504/L23/negative* /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux

#next rename to something easier to work with
for file in ./*
do
   mv "$file" "${file:20}"
done


for file in ./*fastq.genos
do
   mv "$file" "${file%..genos}".genos
done
```


# Genotype  (bad scripts)

Note: this section used gtseq pipeline with a bug.

Here we run an array job to generate .genos files from demuxed fastqs, running 20 files in parallel at a time.

First let's decompress all the files from the latest run
```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-554

#$ -tc 20

#$ -N decompress

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err

FASTQS=(`ls *fastq.gz`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}

gunzip -c $INFILE > ${INFILE%.gz}

#save as script and submit this with qsub -q harold scriptname
```

Also need to do the same for other fastqs: first we'll decompress the 2018 half pounders and move them to a local directory (demux/halfpound_2018)

```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-554

#$ -tc 20

#$ -N decompress

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err

FASTQS=(`##########`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}

gunzip -c $INFILE > ${INFILE%.gz}

#save as script and submit this with qsub -q harold scriptname
```



Next we'll run the GTseq genotyper (v3) script on each fastq file to generate the individual level .genos outputs. This is run for the new data as well as the old. Each set is run as a separate array job. The probe_seq file is full the full 390 SNP panel with allele correction values from CRITFC.

First set the perl environment
```{bash, eval=FALSE}
#installed String::Approx (copied tar.gz and followed install in the readme)

setenv PERL5LIB ~/perl5/lib/perl5/x86_64-linux-thread-multi/ #for tcsh
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/' # for bash

#note i've been running in bash so neet to set the environmental variables a little differently
```

then run the genotyper
first the lane2 data
```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-554

#$ -tc 20

#$ -N GTseq-genotyperv3

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/'

FASTQS=(`ls /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/*fastq`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}
OUTFILE=$(basename ${INFILE%.fastq.gz}.genos)

GTSEQ_GENO="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.pl
"

PROBE_SEQS="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/Omy_GTseq390_ProbeSeqs.csv"

perl $GTSEQ_GENO $PROBE_SEQS $INFILE > $OUTFILE

#save as script and submit this with qsub -q harold scriptname in the ./genos/ dir
```

then the 2018 halfpounders
```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-407

#$ -tc 10

#$ -N GTseq-genotyperv3

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err

#get the 2018 halfpounders
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/'


FASTQS=(`ls /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/halfpound_2018/*fastq`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}

GTSEQ_GENO="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.pl
"

OUTFILE=$(basename ${INFILE%.fastq}.genos)

PROBE_SEQS="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/Omy_GTseq390_ProbeSeqs.csv"

perl $GTSEQ_GENO $PROBE_SEQS $INFILE > $OUTFILE

#save as script and submit this with qsub -q harold scriptname
```


```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-96

#$ -tc 10

#$ -N GTseq-genotyperv3

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err

#get the 2018 summer and winter
FASTQS=(`ls /dfs/FW_HMSC/Omalley_Lab/bohns/GTseq/OmyRogue/baseline/sample_fastqs/*fastq`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}

GTSEQ_GENO="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.pl
"

OUTFILE=$(basename ${INFILE%fastq}.genos)

PROBE_SEQS="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/Omy_GTseq390_ProbeSeqs.csv"

perl $GTSEQ_GENO $PROBE_SEQS $INFILE > $OUTFILE

#save as script and submit this with qsub -q harold scriptname
```

After the genos are written for the panel, we add the sex genotyper
```{bash, eval =FALSE}
#the omysex script is hardcoded to require the fastqs and genos to all be in a single collective directory ... 
#rather than try to fix this hardcoding in the script, I just coped with it and temporarily copied all the fastqs into the .genos driectory then deleted them after i was done

cp /dfs1/FW_HMSC/Omalley_Lab/bohns/GTseq/OmyRogue/baseline/sample_fastqs/*fastq ./
cp /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/halfpound_2018/*fastq ./
cp /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/*fastq ./

#oops it seems like the omysex script doesn't like the naming convention I used lets rename
for file in ./*fastq.genos
do
   mv "$file" "${file%.fastq.genos}".genos
done

#below is the script to run the omysex script

SGE_Batch -q harold -r omysex -c 'perl /dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/OmySEX_test_v3.pl'

# don't forget to remove these dups when done


```

After all the .genos are written we compile them into a single output using the GenoCompile script

```{bash, eval=FALSE}

#this is run from within the .genos directory

SGE_Batch -q harold -r compile -c 'perl /dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_GenoCompile_v3.pl > ../genotypes/half_pounder_GTs_0.1.csv'

```

Run summary figs scripts

```{bash, eval=FALSE}

#this is run from within the .genos directory

SGE_Batch -r figs -q harold -c "python /dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_SummaryFigures_v3_troubleshooting.py '/dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/genos'" 

```

# Filtering (bad scripts)

## Filtering and QC outline 
__QAQC Check:__  
- Check positive and negative controls (for plate flipping, other library prep errors)  
- Check het winter summer controls  
- Check technical replicates  
- remove controls and replicates

__Filtering:__  
- IFI_cutoff=2.5  
- GTperc_cutoff=90 (inds greater than 10% missing data excluded)  
- Missingness (loci) > 20%
- Missingness (loci) > 10% - examine for allele correction issues  
- Remove monomorphic SNPs

__File readme__  
0.1: Raw, unfiltered GTs including all controls and replicates  
0.2: Raw, unfiltered GTs, replicates and controls removed  
0.3: filtered for IFI, GTperc (inds), and Missingness (loci) > 20%  
0.4: filtered for IFI, GTperc (inds), and Missingness (loci) > 20%, and allele correction issues  
1.0: 0.4 + removed monomorphic

## QAQC

### Controls

First let's check that the controls worked well. We will check that negative controls have much fewer reads than average (there may be some on-target reads from otehr samples due to index sequence error)
```{r, warning=FALSE, message=FALSE}
#first I cleaned up the sample names with regex in a text editor - separated adapter info from sample name
genos_0.1 <- read_csv("genotype_data/half_pounder_GTs_0.1_bug.csv")

#lets set a value to mark controls
genos_0.1 <- genos_0.1 %>%
  mutate(control = ifelse(grepl("positive", Sample), "positive", ifelse(grepl("negative", Sample), "negative", "sample")))

ggplot()+geom_histogram(data = genos_0.1, aes(x = `On-Target Reads`, fill= control)) + theme_classic()


```

Looks good, but lets just double check that there isn't a negative with a lot of reads hiding in there and indicating a plate flip:

```{r, message=FALSE, warning=FALSE}
ggplot()+geom_histogram(data = genos_0.1[genos_0.1$control=="negative",], aes(x = `On-Target Reads`)) + theme_classic()

```

Uh-oh a negative control with ~6300 on target reads, but maybe there's just a lot of reads at this adapter/well, lets examine as a portion of total reads, and also the portion GT'd

```{r, warning=FALSE}
ggplot()+geom_histogram(data = genos_0.1, aes(x = `%On-Target`, fill= control)) + theme_classic()
```

Okay, that looks a lot better. The negative control with a lot of reads only has ~1.5% on target, much lower than most samples and positive controls. Positive and negative controls check out. Also of note here is that the positive controls are biased down w/r/t on target proportion, suggesting these samples are getting old. 

### Replicates

Some samples were replicated, let's check for concordance in the genotypes, the pick the sample with better GT success and throw out the duplicate.

```{r, cache=TRUE, warning=FALSE, message=FALSE}
genos_0.11 <- genos_0.1 %>%
  filter(control == "sample") %>%
  filter(Sample != "Winter") %>%
  filter(Sample != "Heterozygous") %>%
  filter(Sample != "Summer")
dups <- genos_0.11[genos_0.11$Sample %in% genos_0.11$Sample[duplicated(genos_0.11$Sample)],]
dups <- dups[order(dups$Sample),]

# woof I don't see a good way around using a nested for loop here
dups_genos <- dups[,8:398]
rep_info <- matrix(ncol=ncol(dups_genos), nrow=nrow(dups_genos)/2)
colnames(rep_info) <- colnames(dups_genos)
for (j in 1:(nrow(dups_genos)/2)) {
for (i in 1:ncol(dups_genos)) {
  rep_info[j,i] <- sum(dups_genos[(j*2)-1,i]==dups_genos[(j*2),i])
}
  }

geno_concordance <- as.data.frame(as.matrix(rep_info)) %>%
  rowMeans()

rep_data <- as.data.frame(cbind(dups[c(1:length(geno_concordance))*2,2], geno_concordance))
ggplot(data=rep_data)+geom_histogram(aes(x=geno_concordance))+theme_classic()

```

There are 50 replicated samples. Mean concordance of genotype across replicates is 87%. Next let's examine whats going on with the samples with very low concordance.

```{r}
#get the bad samples
bad_reps <- genos_0.11[genos_0.11$Sample %in% rep_data[rep_data$geno_concordance<0.50,1],1:7]
bad_reps[order(bad_reps$Sample),]
```

One of the pair just has extremely low %On-Target reads

__Replication Summary__
Replicate samples look good, bad replicates (<50% concordance) seems to be due to one replicate having extremely low GT success. 

Next let's make the 0.2 dataset.
```{r}
genos_0.2 <- genos_0.11 %>%
  group_by(Sample) %>%
  filter(`On-Target Reads` == max(`On-Target Reads`))

#lets merge this dataset with our metadata and do a sanity check
meta_data[!(meta_data$ID %in% genos_0.2$Sample),1:3]
```

There are 976 individuals in the meta_data (i.e. were taken in), but 5 (all winter baseline fish) did not have output fastqs in the raw data directory and were excluded from genotyping...

The 0.2 genotype dataset consists of 971 individuals

### Filtering
- IFI_cutoff=2.5  
- GTperc_cutoff=90 (inds greater than 10% missing data excluded)  
- Missingness (loci) > 20%
- Missingness (loci) > 10% - examine for allele correction issues  
- Remove monomorphic SNPs

Next we move on to filtering individuals and loci on IFI, and missingness. 

First let's take a look at the distribution of these values
```{r, message=FALSE, warning=FALSE}
ggplot(genos_0.2)+geom_histogram(aes(x=IFI))+geom_vline(aes(xintercept= 2.5), color="red")+theme_classic()
ggplot(genos_0.2)+geom_histogram(aes(x=`%GT`))+geom_vline(aes(xintercept= 90), color="red")+theme_classic()

missingness <- (colSums(genos_0.2[,8:398] == "00"))/nrow(genos_0.2)
missing <- as.data.frame(missingness)
missing$marker <- row.names(missing)
ggplot(missing) + geom_histogram(aes(x=missingness))+geom_vline(aes(xintercept= 0.2), color="red")+geom_vline(aes(xintercept= 0.1), color="blue")+theme_classic()+xlab("missingness (loci)")
```

Now let's make the datasets.  
__0.3__
```{r}
genos_0.2 %>%
  filter(IFI >2.5)

genos_0.2 %>%
  filter(`%GT` < 90)

sum(missing$missingness>0.2)

bad_markers <- missing[missing$missingness>0.2, 2]

genos_0.3 <- genos_0.2 %>%
  filter(IFI <2.5) %>%
  filter(`%GT` > 90) %>%
  dplyr::select(-one_of(bad_markers))
  
```

__Filtering log 0.2 -> 0.3:__
4 inds removed with IFI > 2.5  
96 inds removed with genotying success less than 90%  
23 loci removed with > 20% missingness  

Leaves a dataset of 875 individuals and 368 markers.

__0.4__
Next we examine markers with moderately bad genotyping success to attempt to figure out what's going on. 


Could not get the summary_figures script to run, so decided to pull the data myself
The script below will pull the allele count ratios and read counts for all individuals in the pipeline
```{bash, eval = FALSE}
#collect marker info from all the genos files
touch marker_info.txt
for file in ./*genos
do
    awk ' FS="," {print FILENAME,$1,$2,$3,$6,$7,$8}' $file >> marker_info.txt
done

#added headers (ind, marker, a1_count, a2_count, called_geno, a1_corr, a2_corr) and cleaned up with regex in texteditor
```

Let's take a look at the bad markers (10-20% missingness)
```{r}
#get marker names of markers with 0.1 > missingness > 0.2
miss0.1 <- missing[missing$missingness > 0.1,]
miss_mod <- miss0.1[miss0.1$missingness < 0.2, 2]
```

There are 38 markers with moderately poor genotyping success (i.e. less than 20% missingess, but greater than 10% missingness). Previously these types of markers were filtered on an adhoc basis based on the extent of individuals with unclear genotypes (a lot of individuals not falling in the clear cutoffs for GT). Will attempt the same below:

Let's explore look at one of the markers individual. We'll create a similar plot to those generated by the GTseq figures script.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
marker_info <- read_tsv("genotype_data/marker_info_bug.txt")
marker_info$a1_count <- as.numeric(marker_info$a1_count)
marker_info$a2_count <- as.numeric(marker_info$a2_count)

#plot for markers with no allele correction
ggplot(data=marker_info[marker_info$marker=="Chr28_11671116",])+geom_point(aes(a1_count, a2_count, color = called_geno))+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0))+geom_abline(aes(slope = 0.1, intercept=0))+geom_abline(aes(slope = 5, intercept=0))+geom_abline(aes(slope = .2, intercept=0))+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+ coord_fixed(ratio =1 , xlim =c(-2,max( marker_info[marker_info$marker=="Chr28_11671116","a1_count"])), ylim = c(-2,max( marker_info[marker_info$marker=="Chr28_11671116","a2_count"])), expand = FALSE )

ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno))+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0))+geom_abline(aes(slope = 0.1, intercept=0))+geom_abline(aes(slope = 0.5, intercept=0))+geom_abline(aes(slope = 5, intercept=0))+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,100))+ylim(c(0,100))

#wtf is going on here, there are NAs called where there shouldn't be according to the gtseq manuscript
marker_info %>%
  filter(called_geno =="HET") %>%
  mutate(ratio = a1_count/a2_count) %>%
  summarise(min=min(ratio, na.rm = TRUE), max=max(ratio, na.rm = TRUE))

#okay the actual cutoffs used were 0.2 and 2, which are not symmetrical

a <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno),alpha=0.5)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.5, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+ coord_fixed(ratio =1 , xlim =c(-2,max( marker_info[marker_info$marker=="OMS00008","a1_count"])), ylim = c(-2,max( marker_info[marker_info$marker=="OMS00008","a2_count"])), expand = FALSE )+ggtitle("(A) cutoff in data and from pipeline: \n(0.2 - 2)(A1/A2) ")


b <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.5, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,50))+ylim(c(0,50))+ggtitle("(B) cutoff in data and from pipeline: \n(0.2 - 2)(A1/A2)    zoomed") #note these lines are inverse of ratios because slope is y/x, plot is allele1 (x) vs allele2 (y) and "ratio" in the script is allele1/allele 2

c <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.2, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,50))+ylim(c(0,50))+ggtitle("(C) stated cutoff in gtseq manuscript\n (0.2 - 5)(A1/A2)")

d <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.5, intercept=0), color = "blue")+geom_abline(aes(slope = 2, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,50))+ylim(c(0,50))+ggtitle("(D) cutoff from summary figs script\n(0.5 - 2)(A1/A2)")

cowplot::plot_grid(a,b,c,d)

plots <- marker_info %>%
  group_by(marker) %>%
  do(plots=ggplot(data=.)+geom_point(aes(a1_count, a2_count, color = called_geno))+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.2, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+ggtitle(unique(.$marker)))

#plot all "bad markers"
plots$plots[plots$marker %in% miss_mod]

```

# Bug Report

This QC step above (not run in notebook, but run code chunk above to see figures) resulted in the discovery of bug in the gtseq genotyper and gtseq summary figure script. The cutoff's reported in the manuscript were different than those in the scripts, and there was inconsistencies in the values used across scripts in the pipeline.

To fix this we renamed the directories containing the bad outputs (appended _bug) and started from demultiplexed reads again. The corrected script adjusts the allele ratio cutoff for hets from 0.2 - 5, by changing line 210 in the v3 script to the following code chunk. This corrected script is at /dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.1.pl

All notebook sections effected have (bad scripts) appended to their title. Notebook section using the updated script appear below and carry (corrected scripts) if they are duplicated.

```{perl, eval=FALSE}
elsif ($ratio <= 5) {$geno = "$allele1name{$loci}$allele2name{$loci}"; $genoclass = "HET";}
```


# Genotype (corrected scripts)

Can start directly from the genotyper script and skip some of the prep done in the genotype (bad script) section.

Next we'll run the GTseq genotyper (v3.1) script on each fastq file to generate the individual level .genos outputs. This is run for the new data as well as the old. Each set is run as a separate array job. The probe_seq file is full the full 390 SNP panel with allele correction values from CRITFC.

First set the perl environment
```{bash, eval=FALSE}
#installed String::Approx (copied tar.gz and followed install in the readme)

setenv PERL5LIB ~/perl5/lib/perl5/x86_64-linux-thread-multi/ #for tcsh
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/' # for bash

#note i've been running in bash so neet to set the environmental variables a little differently
```

then run the genotyper
first the lane2 data
```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-554

#$ -tc 20

#$ -N GTseq-genotyperv3

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/'

FASTQS=(`ls /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/*fastq`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}
OUTFILE=$(basename ${INFILE%.fastq.gz}.genos)

GTSEQ_GENO="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.1.pl
"

PROBE_SEQS="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/Omy_GTseq390_ProbeSeqs.csv"

perl $GTSEQ_GENO $PROBE_SEQS $INFILE > $OUTFILE

#save as script and submit this with qsub -q harold scriptname in the ./genos/ dir
```

then the 2018 halfpounders
```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-407

#$ -tc 10

#$ -N GTseq-genotyperv3

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err

#get the 2018 halfpounders
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/'


FASTQS=(`ls /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/halfpound_2018/*fastq`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}

GTSEQ_GENO="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.1.pl
"

OUTFILE=$(basename ${INFILE%.fastq}.genos)

PROBE_SEQS="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/Omy_GTseq390_ProbeSeqs.csv"

perl $GTSEQ_GENO $PROBE_SEQS $INFILE > $OUTFILE

#save as script and submit this with qsub -q harold scriptname
```


```{bash, eval=FALSE}

#!/bin/bash
#$ -S /bin/bash

#$ -t 1-96

#$ -tc 10

#$ -N GTseq-genotyperv3

#$ -cwd

#$ -o $JOB_NAME_$TASK_ID.out

#$ -e $JOB_NAME_$TASK_ID.err

#get the 2018 summer and winter
export PERL5LIB='/home/fw/dayand/perl5/lib/perl5/x86_64-linux-thread-multi/'


FASTQS=(`ls /nfs1/FW_HMSC/OMalley_Lab/bohns/GTseq/OmyRogue/baseline/sample_fastqs/*fastq`)
INFILE=${FASTQS[$SGE_TASK_ID -1]}

GTSEQ_GENO="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_Genotyper_v3.1.pl
"

OUTFILE=$(basename ${INFILE%fastq}.genos)

PROBE_SEQS="/dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/Omy_GTseq390_ProbeSeqs.csv"

perl $GTSEQ_GENO $PROBE_SEQS $INFILE > $OUTFILE

#save as script and submit this with qsub -q harold scriptname
```


After the genos are written for the panel, we add the sex genotyper
```{bash, eval =FALSE}
#the omysex script is hardcoded to require the fastqs and genos to all be in a single collective directory ... 
#rather than try to fix this hardcoding in the script, I just coped with it and temporarily copied all the fastqs into the .genos driectory then deleted them after i was done

cp /nfs1/FW_HMSC/OMalley_Lab/bohns/GTseq/OmyRogue/baseline/sample_fastqs/*fastq ./
cp /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/halfpound_2018/*fastq ./
cp /dfs/FW_HMSC/Omalley_Lab/dayan/half_pounder/genotyping/demux/*fastq ./

#oops it seems like the omysex script doesn't like the naming convention I used lets rename
for file in ./*fastq.genos
do
   mv "$file" "${file%.fastq.genos}".genos
done

#below is the script to run the omysex script

SGE_Batch -q harold -r omysex -c 'perl /dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/OmySEX_test_v3.pl'

# don't forget to remove these dups when done


```

After all the .genos are written we compile them into a single output using the GenoCompile script

```{bash, eval=FALSE}

#this is run from within the .genos directory

SGE_Batch -q harold -r compile -c 'perl /dfs/FW_HMSC/Omalley_Lab/dayan/software/GTseq-Pipeline/GTseq_GenoCompile_v3.pl > ../genotypes/half_pounder_GTs_0.1.csv'

```

# Filtering (corrected scripts)

## Filtering and QC outline 
__QAQC Check:__  
- Check positive and negative controls (for plate flipping, other library prep errors)  
- Check het winter summer controls  
- Check technical replicates  
- remove controls and replicates

__Filtering:__  
- IFI_cutoff=2.5  
- GTperc_cutoff=90 (inds greater than 10% missing data excluded)  
- Missingness (loci) > 20%
- Missingness (loci) > 10% - examine for allele correction issues  
- Remove monomorphic SNPs
- Remove duplicated individuals

__File readme__  
0.1: Raw, unfiltered GTs including all controls and replicates  
0.2: Raw, unfiltered GTs, replicates and controls removed  
0.3: filtered for IFI, GTperc (inds), and Missingness (loci) > 20%  
0.4: filtered for IFI, GTperc (inds), and Missingness (loci) > 20%, and allele correction issues  
1.0: 0.4 + removed monomorphic  
2.0: 1.0 + duplicate sample removal  
2.2:  final genotype dataset with metadata (population, run, sample date)  

## QAQC

### Controls

First let's check that the controls worked well. We will check that negative controls have much fewer reads than average (there may be some on-target reads from otehr samples due to index sequence error)
```{r, warning=FALSE, message=FALSE}
#first I cleaned up the sample names with regex in a text editor - separated adapter info from sample name
genos_0.1 <- read_csv("genotype_data/half_pounder_GTs_0.1.csv")

#lets set a value to mark controls
genos_0.1 <- genos_0.1 %>%
  mutate(control = ifelse(grepl("positive", Sample), "positive", ifelse(grepl("negative", Sample), "negative", "sample")))

ggplot()+geom_histogram(data = genos_0.1, aes(x = `On-Target Reads`, fill= control)) + theme_classic()


```

Looks good, but lets just double check that there isn't a negative with a lot of reads hiding in there and indicating a plate flip:

```{r, message=FALSE, warning=FALSE}
ggplot()+geom_histogram(data = genos_0.1[genos_0.1$control=="negative",], aes(x = `On-Target Reads`)) + theme_classic()

```

Uh-oh a negative control with ~6300 on target reads, but maybe there's just a lot of reads at this adapter/well, lets examine as a portion of total reads, and also the portion GT'd

```{r, warning=FALSE}
ggplot()+geom_histogram(data = genos_0.1, aes(x = `%On-Target`, fill= control)) + theme_classic()
```

Okay, that looks a lot better. The negative control with a lot of reads only has ~XXXXX % on target, much lower than most samples and positive controls. Positive and negative controls check out. Also of note here is that the positive controls are biased down w/r/t on target proportion, suggesting these samples are getting old. 

### Replicates

Some samples were replicated, let's check for concordance in the genotypes, the pick the sample with better GT success and throw out the duplicate.

```{r, cache=TRUE, warning=FALSE, message=FALSE}
#change to undo cache
genos_0.11 <- genos_0.1 %>%
  filter(control == "sample") %>%
  filter(Sample != "Winter") %>%
  filter(Sample != "Heterozygous") %>%
  filter(Sample != "Summer")
dups <- genos_0.11[genos_0.11$Sample %in% genos_0.11$Sample[duplicated(genos_0.11$Sample)],]
dups <- dups[order(dups$Sample),]

# woof I don't see a good way around using a nested for loop here
dups_genos <- dups[,8:398]
rep_info <- matrix(ncol=ncol(dups_genos), nrow=nrow(dups_genos)/2)
colnames(rep_info) <- colnames(dups_genos)
for (j in 1:(nrow(dups_genos)/2)) {
for (i in 1:ncol(dups_genos)) {
  rep_info[j,i] <- sum(dups_genos[(j*2)-1,i]==dups_genos[(j*2),i])
}
  }

geno_concordance <- as.data.frame(as.matrix(rep_info)) %>%
  rowMeans()

rep_data <- as.data.frame(cbind(dups[c(1:length(geno_concordance))*2,2], geno_concordance))
ggplot(data=rep_data)+geom_histogram(aes(x=geno_concordance))+theme_classic()

```

There are 50 replicated samples. Mean concordance of genotype across replicates is 87.3%. Next let's examine whats going on with the samples with very low concordance.

```{r}
#get the bad samples
bad_reps <- genos_0.11[genos_0.11$Sample %in% rep_data[rep_data$geno_concordance<0.50,1],1:7]
bad_reps[order(bad_reps$Sample),]
```

One of the pair just has extremely low %On-Target reads

__Replication Summary__
Replicate samples look good, bad replicates (<50% concordance) seems to be due to one replicate of the pair having extremely low GT success. 

Next let's make the 0.2 dataset.
```{r}
genos_0.2 <- genos_0.11 %>%
  group_by(Sample) %>%
  filter(`On-Target Reads` == max(`On-Target Reads`))

#lets merge this dataset with our metadata and do a sanity check
meta_data[!(meta_data$ID %in% genos_0.2$Sample),1:3]
```

There are 976 individuals in the meta_data (i.e. were taken in), but 8 (all baseline fish) did not have output fastqs in the raw data directory and were excluded from genotyping...

The 0.2 genotype dataset consists of 971 individuals

### Filtering

- IFI_cutoff=2.5  
- GTperc_cutoff=90 (inds greater than 10% missing data excluded)  
- Missingness (loci) > 20%
- Missingness (loci) > 10% - examine for allele correction issues  
- Remove monomorphic SNPs

Next we move on to filtering individuals and loci on IFI, and missingness. 

First let's take a look at the distribution of these values
```{r, message=FALSE, warning=FALSE}
ggplot(genos_0.2)+geom_histogram(aes(x=IFI))+geom_vline(aes(xintercept= 2.5), color="red")+theme_classic()
ggplot(genos_0.2)+geom_histogram(aes(x=`%GT`))+geom_vline(aes(xintercept= 90), color="red")+theme_classic()

missingness <- (colSums(genos_0.2[,8:398] == "00"))/nrow(genos_0.2)
missing <- as.data.frame(missingness)
missing$marker <- row.names(missing)
ggplot(missing) + geom_histogram(aes(x=missingness))+geom_vline(aes(xintercept= 0.2), color="red")+geom_vline(aes(xintercept= 0.1), color="blue")+theme_classic()+xlab("missingness (loci)")
```

Now let's make the datasets.  
__0.3__
```{r}
genos_0.2 %>%
  filter(IFI >2.5)

genos_0.2 %>%
  filter(`%GT` < 90)

sum(missing$missingness>0.2)

bad_markers <- missing[missing$missingness>0.2, 2]

genos_0.3 <- genos_0.2 %>%
  filter(IFI <2.5) %>%
  filter(`%GT` > 90) %>%
  dplyr::select(-one_of(bad_markers))
  
```

__Filtering log 0.2 -> 0.3:__  
5 inds removed with IFI > 2.5  
49 inds removed with genotying success less than 90%  
19 loci removed with > 20% missingness  

Leaves a dataset of 918 individuals and 373 markers.  

__0.4__
Next we examine markers with moderately bad genotyping success to attempt to figure out what's going on. 


Could not get the summary_figures script to run, so decided to pull the data myself
The script below will pull the allele count ratios and read counts for all individuals in the pipeline
```{bash, eval = FALSE}
#collect marker info from all the genos files
touch marker_info.txt
for file in ./*genos
do
    awk ' FS="," {print FILENAME,$1,$2,$3,$6,$7,$8}' $file >> marker_info.txt
done

#added headers (ind, marker, a1_count, a2_count, called_geno, a1_corr, a2_corr) and cleaned up with regex in texteditor
```

Let's take a look at the bad markers (10-20% missingness)
```{r}
#get marker names of markers with 0.1 > missingness > 0.2
miss0.1 <- missing[missing$missingness > 0.1,]
miss_mod <- miss0.1[miss0.1$missingness < 0.2, 2]
```

There are 18 markers with moderately poor genotyping success (i.e. less than 20% missingess, but greater than 10% missingness)(an improvement of 20 markers from the genotyper script with an error). Previously these types of markers were filtered on an adhoc basis based on the extent of individuals with unclear genotypes (a lot of individuals not falling in the clear cutoffs for GT). Will attempt the same below:

Let's explore look at one of the markers individual. We'll create a similar plot to those generated by the GTseq figures script.

Below we show plots for one representative good marker, and one representative marker with issues. 
```{r, message=FALSE, warning=FALSE}
marker_info <- read_tsv("genotype_data/marker_info.txt")
marker_info$a1_count <- as.numeric(marker_info$a1_count)
marker_info$a2_count <- as.numeric(marker_info$a2_count)

#plot for markers with no allele correction
#ggplot(data=marker_info[marker_info$marker=="Chr28_11671116",])+geom_point(aes(a1_count, a2_count, color = called_geno))+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0))+geom_abline(aes(slope = 0.1, intercept=0))+geom_abline(aes(slope = 5, intercept=0))+geom_abline(aes(slope = .2, intercept=0))+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+ coord_fixed(ratio =1 , xlim =c(-2,max( marker_info[marker_info$marker=="Chr28_11671116","a1_count"])), ylim = c(-2,max( marker_info[marker_info$marker=="Chr28_11671116","a2_count"])), expand = FALSE )

#ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno))+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0))+geom_abline(aes(slope = 0.1, intercept=0))+geom_abline(aes(slope = 0.5, intercept=0))+geom_abline(aes(slope = 5, intercept=0))+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,100))+ylim(c(0,100))

#wtf is going on here, there are NAs called where there shouldn't be according to the gtseq manuscript
marker_info %>%
  filter(called_geno =="HET") %>%
  mutate(ratio = a1_count/a2_count) %>%
  summarise(min=min(ratio, na.rm = TRUE), max=max(ratio, na.rm = TRUE))

#okay the actual cutoffs used were 0.2 and 2, which are not symmetrical

# a <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno),alpha=0.5)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.5, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+ coord_fixed(ratio =1 , xlim =c(-2,max( marker_info[marker_info$marker=="OMS00008","a1_count"])), ylim = c(-2,max( marker_info[marker_info$marker=="OMS00008","a2_count"])), expand = FALSE )+ggtitle("(A) cutoff in data and from pipeline: \n(0.2 - 2)(A1/A2) ")


# b <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.5, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,50))+ylim(c(0,50))+ggtitle("(B) cutoff in data and from pipeline: \n(0.2 - 2)(A1/A2)    zoomed") #note these lines are inverse of ratios because slope is y/x, plot is allele1 (x) vs allele2 (y) and "ratio" in the script is allele1/allele 2

ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.2, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,150))+ylim(c(0,150))+ggtitle("(C) stated cutoff in gtseq manuscript\n (0.2 - 5)(A1/A2)")

ggplot(data=marker_info[marker_info$marker=="Omy_RAD46672-27",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.2, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,150))+ylim(c(0,150))+ggtitle("(C) stated cutoff in gtseq manuscript\n (0.2 - 5)(A1/A2)")

#d <- ggplot(data=marker_info[marker_info$marker=="OMS00008",])+geom_point(aes(a1_count, a2_count, color = called_geno), alpha = 0.7, size =2)+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.5, intercept=0), color = "blue")+geom_abline(aes(slope = 2, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+xlim(c(0,50))+ylim(c(0,50))+ggtitle("(D) cutoff from summary figs script\n(0.5 - 2)(A1/A2)")


plots <- marker_info %>%
  group_by(marker) %>%
  do(plots=ggplot(data=.)+geom_point(aes(a1_count, a2_count, color = called_geno))+theme_classic()+geom_abline(aes(slope=1, intercept=0))+geom_abline(aes(slope = 10, intercept=0), color = "green")+geom_abline(aes(slope = 0.1, intercept=0), color = "red")+geom_abline(aes(slope = 0.2, intercept=0), color = "blue")+geom_abline(aes(slope = 5, intercept=0), color = "blue")+coord_equal(ratio=1)+geom_abline(slope = -1, intercept = 10)+ggtitle(unique(.$marker)))

#plot all "bad markers"
#mod_bad_plot_index <- which(plots$marker %in% miss_mod)
# plots$plots[[mod_bad_plot_index[10]]] #manually looped through these plots by changing the index for all 28 moderately bad markers, could make an lapply loop in the future, bad markers reported below

```

Among the markers with moderately bad genotyping success the following 3 markers had issues and were removed. 

Omy_RAD46672-27 - a2 biased  
Omy_RAD52458-17 - a1 biased line (and 1:1), population level variation in paralog divergence?


```{r}
to_filt <- c("Omy_RAD46672-27", "Omy_RAD52458-17")
genos_0.4 <- genos_0.3 %>%
  dplyr::select(-one_of(to_filt))
```

Leaves a dataset of 918 individuals and 370 markers.

__1.0__

To generate the 1.0 dataset, we remove monomorphic markers

```{r}
genos_1.0 <- genos_0.4 %>% 
  select_if(~ length(unique(.)) > 1)
```

This removed 16 monomorphic loci. The 1.0 dataset has 918 individuals and 355 markers. 

__Duplicate Samples__

Some sample tissues were provided in batches of ~50 fin clips. Let's make sure no fin clips broke apart leading to a single individual to be represented twice in the dataset. Rather than fussing with installing coancestry for windows on a unix system, estimated relatedness using an R package (related) which can implement the code from Coancester

We used the estimator from Lynch and Ritland 1999 #not dyadic likelihood estimator, Milligan (2003) 
```{r, eval=FALSE}
# needs a unique row for each indiviudal and two columns for each diploid locus
# threw out metadata nd wrote to a file
# then we split all the genotype values using regex in a text editor (after converting all na values to 0)
#   find string: \t([ATGC0XY])([ATCG0XY])  replace string: \t\1\t\2
# convert genos to numbers and removed sex marker
#convert to integer T-1 G->2 etc

just_genos <- genos_1.0[,c(2, 8:357)]
write_tsv(just_genos, "genotype_data/just_genos.txt")

# now run coancestry
rmat <- coancestry("./genotype_data/just_genos.txt", dyadml = 1)
rmat2 <- coancestry("./genotype_data/just_genos.txt", lynchrd  = 1)

# save the relevant info so we don't have to run this over and over and take up a ton of diskspace
rmat_to_save <- rmat2$relatedness[rmat2$relatedness$lynchrd > 0.5,]
save(rmat_to_save, file="genotype_data/relatedness.Rdata")
```

Check for highly related individuals and remove any >= 1 from the dataset
```{r}
#Check for relatedness
load(file = "genotype_data/relatedness.Rdata")
#ggplot(rmat_to_save$relatedness)+geom_histogram(aes(x=lynchrd))+theme_classic()
rmat_to_save[which(rmat_to_save$lynchrd >=0.95), c(1:3)]

dup_inds <- rmat_to_save[which(rmat_to_save$lynchrd >= 0.95), c(1:3)]

genos_2.0 <- genos_1.0 %>%
  filter(!(Sample %in% dup_inds$ind2.id))
```

There were 39 pairs of duplicated samples. All (but two pairs) were among the half pounders and within a single year. All were also within <50 of each other w respct to their ID number, suggesting they were from the same jar. 

The final filtered dataset contains 882 samples genotyped at 355 markers.

# File Conversion and Stats

Final step of genotyping is to collect some stats about the genotype dataset and reformat the genotype file into common formats for import.

## Stats


```{r, fig.cap="On Target Read Distribution"}
ggplot(genos_2.0)+geom_density(aes(x=`On-Target Reads`))+geom_vline(aes(xintercept=median(`On-Target Reads`)), color = "red") +theme_classic()
```

The median number of on target reads was 98104. There were 350,847,984 reads contributing to the samples included in the final dataset and 407,683,506 reads contributing to all individuals. 

```{r, fig.cap="Proportion on Target"}
ggplot(genos_2.0)+geom_density(aes(x=`%On-Target`))+geom_vline(aes(xintercept=median(`%On-Target`)), color = "red") +theme_classic()
```

The median proportion of on target reads was 30%.

```{r, warning=FALSE, message=FALSE}
#code to estimate depth at filtered loci
marker_info %>%
  filter(marker %in% colnames(genos_2.0)) %>%
  mutate(sumdepth=a1_count+a2_count) %>%
  summarise(mean=mean(sumdepth, na.rm = TRUE), median=median(sumdepth, na.rm = TRUE), sd=sd(sumdepth, na.rm = TRUE))

marker_info %>%
  filter(marker %in% colnames(genos_2.0)) %>%
  mutate(sumdepth=a1_count+a2_count) %>%
  ggplot + aes(x=sumdepth)+geom_histogram()+theme_classic()
```

The median sequencing depth was 166, and the mean was 278 +- 375 (sd)

A note here, this is a huge depth with a huge variance, optimizing primers to reduce variance could allow many fold increase in multiplexing for a lane if desired. Also there is a huge right tail, suggesting some primers are hitting highly repetive sequence and are hogging sequencing.

Finally here are the number of samples per run after filtering
```{r}
genos_2.0 %>%
  left_join(meta_data, by = c("Sample" = "ID")) %>%
  group_by(run, year) %>%
  tally()
```


## Conversion

Let's get some usable file formats

```{r, eval=FALSE}

# import into adegenet

#first get a matrix to work on

#first change column to not include a dot
genos_2.1 <- genos_2.0
colnames(genos_2.1) <- gsub("\\.", "_", colnames(genos_2.1))
#convert to matrix with inds as row names
genos_2.1 <- as.matrix(genos_2.1[,c(8:362)])
row.names(genos_2.1) <- genos_2.0$Sample
genind_1.0 <- df2genind(genos_2.1, sep ="", ploidy=2,NA.char = "0")

#add in the populations
genos_2.2 <- genos_2.0 %>%
  left_join(meta_data, by=c("Sample" = "ID"))

genind_1.0@pop <- as.factor(genos_2.2$run)

```

```{r, eval = FALSE}
# here we save a few objects with useful info
genind_2.0 <- genind_1.0
save(genos_2.2, file ="./genotype_data/genotypes_2.2.R")
save(genind_2.0, file= "./genotype_data/genind_2.0.R")
```

